---
title: "nonhierarchical_stenella_AM"
author: "Raija Hammad"
format: html
editor: visual
---

## Objective

This analysis uses non-hierarchical Bayesian logistic regression to estimate age at maturity for three *Stenella* species: *S. attenuata, S. coerueleoalba,* and *S. longirostris*.

Things to note:

- Maturity status is modeled as a binary response as a function of age for each species.

- The age corresponding to a 50% probability of maturity is used as my estimate of age at maturity.

## Loading Packages

This analysis uses tidyverse for manipulating the data, coda for posterior samples, and rjags/jagsUI to run Bayesian models in JAGS.

```{r}
#| echo: false
library(tidyverse)
library(coda)
library(rjags)
library(jagsUI)

save.output <- TRUE

source("loading_data.R")
```

## Data Preparation for Age and Maturity

Observations with both age and maturity status were kept; rows containing missing values were removed.

#### *Stenella attenuata*

```{r}
AM.data.ATT <- attenuata %>%
  select(Age, isMature_bin) %>%
  drop_na()

mean.age.ATT <- mean(AM.data.ATT$Age)
sd.age.ATT <- sd(AM.data.ATT$Age)
```

#### *Stenella coerueleoalba*

```{r}
AM.data.COE <- coeruleoalba %>%
  select(Age, isMature_bin) %>%
  drop_na()

mean.age.COE <- mean(AM.data.COE$Age)
sd.age.COE <- sd(AM.data.COE$Age)
```

#### *Stenella longirostris*

```{r}
AM.data.LON <- longirostris %>%
  select(Age, isMature_bin) %>%
  drop_na()

mean.age.LON <- mean(AM.data.LON$Age)
sd.age.LON <- sd(AM.data.LON$Age)
```

The resulting datasets contains individuals with usable age and maturity information.

## Bayesian Logistic Regression

Naming constants and vectors in the data for JAGS. For each individual \\(i\\), maturity status is treated as a Bernoulli response.

$$y_i \sim \text{Bernoulli}(p_i)$$

with maturity probability related to age through the logit function:

$$\text{logit}(p_i) = \alpha + \beta X_i$$

where \\(X_i\\) is age, \\(\\alpha\\) is the intercept, and \\(\\beta\\) describes the relationship between age and probability of maturity.

#### *Stenella attenuata*

```{r}
jags.data.AM.ATT <- list(
  N = nrow(AM.data.ATT),
  X = as.vector(AM.data.ATT$Age),
  y = AM.data.ATT$isMature_bin
)
```

#### *Stenella Coerueleoalba*

```{r}
jags.data.AM.COE <- list(
  N = nrow(AM.data.COE),
  X = as.vector(AM.data.COE$Age),
  y = AM.data.COE$isMature_bin
)
```

#### *Stenella Longirostris*

```{r}
jags.data.AM.LON <- list(
  N = nrow(AM.data.LON),
  X = as.vector(AM.data.LON$Age),
  y = AM.data.LON$isMature_bin
)
```

### MCMC

Five independent MCMC chains were run, each used 50000 iterations with the first 25000 discarded as burn-in and every 100th posterior sample kept.

#### *Stenella attenuata*

```{r}
MCMC.AM.ATT <- list(
  n.samples = 50000,
  n.thin = 100,
  n.burnin = 25000,
  n.chains = 5
)

jags.params.AM.ATT <- c("alpha", "beta")
```

#### *Stenella Coerueleoalba*

```{r}
MCMC.AM.COE <- list(
  n.samples = 50000,
  n.thin = 100, 
  n.burnin = 25000,
  n.chains = 5
)

jags.params.AM.COE <- c("alpha", "beta")
```

#### *Stenella Longirostris*

```{r}
MCMC.AM.LON <- list(
  n.samples = 50000,
  n.thin = 100, 
  n.burnin = 25000,
  n.chains = 5
)

jags.params.AM.LON <- c("alpha", "beta")
```

### Model

#### *Stenella attenuata*

```{r}
if (!file.exists("RData/NH.AM.ATT.rds")){
NH.AM.ATT <- jagsUI::jags(
  data = jags.data.AM.ATT,
  inits = NULL,
  parameters.to.save = jags.params.AM.ATT,
  model.file = "models/model_logistic_regression_1sp.jags",
  n.chains = MCMC.AM.ATT$n.chains,
  n.burnin = MCMC.AM.ATT$n.burnin,
  n.thin = MCMC.AM.ATT$n.thin,
  n.iter = MCMC.AM.ATT$n.samples,
  DIC = TRUE,
  parallel = TRUE
)

if (save.output)
  saveRDS(NH.AM.ATT,
          file = "RData/NH.AM.ATT.rds")
  
} else {
  NH.AM.ATT <- readRDS("RData/NH.AM.ATT.rds")
}
```

#### *Stenella Coerueleoalba*

```{r}
NH.AM.COE <- jagsUI::jags(
  data = jags.data.AM.COE,
  inits = NULL,
  parameters.to.save = jags.params.AM.COE,
  model.file = "models/model_logistic_regression_1sp.jags",
  n.chains = MCMC.AM.COE$n.chains,
  n.burnin = MCMC.AM.COE$n.burnin,
  n.thin = MCMC.AM.COE$n.thin,
  n.iter = MCMC.AM.COE$n.samples,
  DIC = TRUE,
  parallel = TRUE
)

if (save.output)
  saveRDS(NH.AM.COE,
          file = "RData/NH.AM.COE")
```

#### *Stenella Longirostris*

```{r}
NH.AM.LON <- jagsUI::jags(
  data = jags.data.AM.LON,
  inits = NULL,
  parameters.to.save = jags.params.AM.LON,
  model.file = "models/model_logistic_regression_1sp.jags",
  n.chains = MCMC.AM.LON$n.chains,
  n.burnin = MCMC.AM.LON$n.burnin,
  n.thin = MCMC.AM.LON$n.thin,
  n.iter = MCMC.AM.LON$n.samples,
  DIC = TRUE,
  parallel = TRUE
)

if (save.output)
  saveRDS(NH.AM.LON,
          file = "RData/NH.AM.LON")
```

# Results

### Extracting Posterior Samples

#### *Stenella attenuata*

```{r}
all.samples.ATT <- as.matrix(NH.AM.ATT$samples)

alpha.samples.ATT <- all.samples.ATT[, "alpha"]
beta.samples.ATT <- all.samples.ATT[, "beta"]
```

#### *Stenella Coerueleoalba*

```{r}
all.samples.COE <- as.matrix(NH.AM.COE$samples)

alpha.samples.COE <- all.samples.COE[, "alpha"]
beta.samples.COE <- all.samples.COE[, "beta"]
```

#### *Stenella Longirostris*

```{r}
all.samples.LON <- as.matrix(NH.AM.LON$samples)

alpha.samples.LON <- all.samples.LON[, "alpha"]
beta.samples.LON <- all.samples.LON[, "beta"]
```

### Age Estimate

The age at which an animal has a 50% probability of being mature.

#### *Stenella attenuata*

```{r}
p.50.AM.ATT <- (-alpha.samples.ATT / beta.samples.ATT)
```

#### *Stenella Coerueleoalba*

```{r}
p.50.AM.COE <- (-alpha.samples.COE / beta.samples.COE)
```

#### *Stenella Longirostris*

```{r}
p.50.AM.LON <- (-alpha.samples.LON / beta.samples.LON)
```

### Summarizing Posterior Estimates

#### *Stenella attenuata*

```{r}
summary(p.50.AM.ATT)

results.AM.ATT <- quantile( 
  p.50.AM.ATT, 
  probs = c(0.025, 0.5, 0.975)
)
```
S. attenuata's age at maturity was `r (results.AM.ATT[2])` (median) with 95% CI of `r signif(results.AM.ATT[1],3)` - `r signif(results.AM.ATT[3],3)`. 



#### *Stenella Coerueleoalba*

```{r}
summary(p.50.AM.COE)

quantile(
  p.50.AM.COE,
  probs = c(0.025, 0.5, 0.975)
)
```

#### *Stenella Longirostris*

```{r}
summary(p.50.AM.LON)

quantile(
  p.50.AM.LON,
  probs = c(0.025, 0.5, 0.975)
)
```
